---
title: "DV_R-Assignment (FINAL)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1

1.  [**Data Inspection:**]{.ul}

-   To set working directory and open files:

    ```{r}
      setwd("~/Desktop/DV_R-Assignment")
      genotypes <- read.delim("fang_et_al_genotypes.txt")
      snp_pos <-read.delim("snp_position.txt")
    ```

1.  To inspect the files:

    -   `file.info` = is used to determine size of file in bytes, whether the file is a directory, last status change and last access times, user ID of the file's owner, username, group name, and other info.

    -   `file.size` = to determine file size in bytes

    -   `dim` = to determine number of rows and columns

        ```{r, results='hide'}
          
        summary(genotypes)
        summary(snp_pos)

        file.info("fang_et_al_genotypes.txt") 
        file.info("snp_position.txt") 

        file.size("fang_et_al_genotypes.txt") 
        file.size("snp_position.txt") 

        dim(genotypes) 
        dim(snp_pos)
        ```

2.  [**Data processing:**]{.ul}

-   First we need to install and run the following packages:

    ```{r}

    if(!require("tibble")) install.packages("tibble")
    library(tibble)

    if(!require("dplyr")) install.packages("dplyr")
    library(dplyr)

    ```

-   Next, need to parse the data out for groups:

    -   Maize: ZMMIL, ZMMLR, and ZMMMR

    -   Teosinte: ZMPBA, ZMPIL, and ZMPJA

        ```{r,}
        maize_genotypes <- subset(genotypes, Group == "ZMMIL" | Group == "ZMMLR" | Group == "ZMMMR")

        teo_genotypes <- subset (genotypes, Group == "ZMPBA" | Group == "ZMPIL" | Group == "ZMPJA")

        ```

-   Third, we need to transpose the data so that our columns in the `fang_et_al_genotypes.txt` file becomes our rows:

    ```{r}
    transposed_maize <- as.data.frame(t(maize_genotypes))
    transposed_teo <- as.data.frame(t(teo_genotypes))

    #Progress check:
    View(transposed_maize)
    View(transposed_teo)

    ```

-   When looking at the data, column one doesn't have name - need to rename it as "SNP_ID":

    ```{r}

    names(transposed_maize) <- lapply(transposed_maize[1,], as.character) #this makes column one a character 

    transposed_maize <- transposed_maize [-1,] 

    names(transposed_teo) <- lapply(transposed_teo[1, ], as.character)
    transposed_teo <- transposed_teo[-1,]

    new_transposed_maize <- rownames_to_column(transposed_maize, var="SNP_ID")
    new_transposed_teo <- rownames_to_column(transposed_teo, var="SNP_ID")
    ```

-   Next, sort transposed_maize and transposed_teosinte by SNP_ID:

    ```{r}
    sorted_maize <- arrange(new_transposed_maize, SNP_ID)
    sorted_teo<- arrange(new_transposed_teo, SNP_ID) 
    ```

-   Parse the snp data to only have SNP_ID, chromsome\_ and position and sort by SNP_ID:

    ```{r}
    snp_pos_chrom <- snp_pos[, c("SNP_ID", "Chromosome", "Position")]
    sorted_snp_pos_chrom <- arrange(snp_pos_chrom, SNP_ID)
    ```

-   Join sorted_maize genotypes and sorted_teosinte with the sorted_snp_pos_chrom data:

    ```{r}
    joined_maize <- merge(sorted_snp_pos_chrom,sorted_maize, by.x="SNP_ID", by.y="SNP_ID", all = TRUE)
    joined_teosinte <- merge(sorted_snp_pos_chrom,sorted_teo, by.x="SNP_ID", by.y="SNP_ID", all = TRUE)

    #Progress check:
    View(joined_teosinte)
    View(joined_maize)
    ```

-   Extract data from each chromosome in maize and teosinte with increasing position with missing data indicated as "?":

    ```{r}

    #sort increasing first: 

    increase_pos_maize <- joined_maize[order(as.numeric(as.character(joined_maize$Position))),]

    increase_pos_teo <- joined_teosinte[order(as.numeric(as.character(joined_teosinte$Position))),]

    #set directory to ensure your files are saved where you want them to: 
    setwd("~/Desktop/DV_R-Assignment/All_Maize_Genotypes/Increasing")

    for (i in 1:10) {
      maize_files_inc <- increase_pos_maize[increase_pos_maize$Chromosome == i,]
      #we can create a new CSV file for each chromosome
      write.csv(maize_files_inc, sprintf("Chr_%d_maize_increasing",i),row.names = F)
    }

    setwd("~/Desktop/DV_R-Assignment/All_Teosinte_Genotypes/Increasing")
    for (i in 1:10){
      teosinte_files_inc<-increase_pos_teo[increase_pos_teo$Chromosome==i,]
      write.csv(teosinte_files_inc,sprintf("Chr_%d_teosinte_increasing",i),row.names = F)
    }
    ```

-   Extract data from each chromosome in maize and teosinte with decreasing position with missing data indicated as "-":

    ```{r}
    #sort decreasing first:
    decrease_pos_maize <- joined_maize[order(-as.numeric(as.character(joined_maize$Position))),]
    decrease_pos_teo <- joined_teosinte[order(-as.numeric(as.character(joined_teosinte$Position))),]

    #replace missing values as "-" 
    decrease_pos_maize[decrease_pos_maize == "?/?"] <- "-"
    decrease_pos_teo [decrease_pos_teo == "?/?"] <- "-"

    #check progress
    View(decrease_pos_maize)
    View(decrease_pos_teo)

    #set directory to ensure your files are saved where you want them to
    setwd("~/Desktop/DV_R-Assignment/All_Maize_Genotypes/Decreasing")

    for (i in 1:10) {
      maize_files_dec <- decrease_pos_maize[decrease_pos_maize$Chromosome == i,]
      #we can create a new CSV file for each chromosome
      write.csv(maize_files_dec, sprintf("Chr_%d_maize_decreasing",i),row.names = F)
    }

    setwd("~/Desktop/DV_R-Assignment/All_Teosinte_Genotypes/Decreasing")
    for (i in 1:10){
      teosinte_files_decrease<-decrease_pos_teo[decrease_pos_teo$Chromosome==i,]
      write.csv(teosinte_files_decrease,sprintf("Chr_%d_teosinte_Decreasing",i),row.names = F)
    }
    ```

## Part 2

1.  [**Graphing Data - SNPs Per Chromosome:**]{.ul}

-   Load the following packages:

    ```{r}
    #load the following 
     if (!require("ggplot2")) install.packages("ggplot2")
    library(ggplot2)
    if(!require("tibble")) install.packages("tibble")
    library(tibble)
    if (!require("reshape2")) install.packages("reshape2")
    library(reshape2)
    if(!require("dplyr")) install.packages("dplyr")
    library(dplyr)
    if (!require("plyr")) install.packages("plyr")
    library(plyr)
    ```

-   Transpose the original genotypes data frame:

    ```{r}
    transposed_genotypes <- as.data.frame(t(genotypes))

    names(transposed_genotypes) <- lapply(transposed_genotypes[1,], as.character) #this makes column one a character 

    transposed_genotypes <- transposed_genotypes[-1,] 

    new_transposed_genotypes <- rownames_to_column(transposed_genotypes, var="SNP_ID")
    ```

-   Sort new_transposed_genotypes with SNP_ID and join with sorted_snp_pos_chrom:

    ```{r}
    sorted_genotypes <- arrange(new_transposed_genotypes, SNP_ID)

    joined_genotypes <- merge(sorted_snp_pos_chrom,sorted_genotypes, by.x="SNP_ID", by.y="SNP_ID", all = TRUE)
    ```

-   Graphing SNPs Per Chromosome:

    ```{r}
    ggplot(data = joined_genotypes) +
      geom_bar(mapping = aes(x = Chromosome)) +
      scale_x_discrete(limits=c(1:10, "unknown", "multiple")) +
      ggtitle(label = "SNPs per chromosome") +
      xlab(label = "Chromosome #") +
      ylab(label = "# of SNPs") +
      theme(
        
        plot.title = element_text(hjust = 0.5, size = 16),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12)
      )

    ```

2.  [**Graphing Data - Missing Data Heterozygosity:**]{.ul}

-   Reorganize the data:

    ```{r}

    SNP_header <- colnames(genotypes)[-c(1:3)] #making a header for all SNP_IDs. 
    m.genotype  <- melt(genotypes, measure.vars = SNP_header) #using melt function make our SNP_ID as rows

    colnames(m.genotype)[4] <- "SNP_ID" #renaming the 4th column from "variable" to SNP_ID

    m.genotype[ m.genotype == "?/?" ] = "N/A"

    ```

-   Use `mutate()` to create new df with new column called "HOMOZYGOUS" with TRUE and FALSE:

    ```{r}
    m.genotype_new <- 
      mutate(m.genotype, Homozygous = m.genotype$value=="A/A" | m.genotype$value=="C/C" | m.genotype$value=="G/G" | m.genotype$value=="T/T")

    #replace th TRUE and FALSE with "Homozygous", "Heterozygous" and "Missing"
    m.genotype_new$Homozygous [ m.genotype_new$Homozygous == "TRUE"] = "Homozygous"
    m.genotype_new$Homozygous [ m.genotype_new$Homozygous == "FALSE"] = "Heterozygous"
    m.genotype_new$Homozygous [ m.genotype_new$value == "N/A"] = "Missing"

    #progress report
    view(m.genotype_new)
    head(m.genotype_new) 
    ```

-   Plot by Sample_ID:

    ```{r}
    ggplot(data = m.genotype_new) +
      geom_bar(mapping = aes(x=Sample_ID, fill=Homozygous)) +
      scale_x_discrete(labels=NULL)+
      ggtitle(label = "Heterozygosity By Sample ID") +
      xlab(label = "Sample ID") +
      ylab(label = "Amount")
    ```

-   Plot by Group:

    ```{r}
    ggplot(data = m.genotype_new) +
      geom_bar(mapping = aes(x=Group, fill=Homozygous), position="dodge")
    ```

3.  [**Graphing Data - Create your own plot: Amount of heterozygosity in each chromosome**]{.ul}

    ```{r}
    m.joined_geno <- merge(sorted_snp_pos_chrom, m.genotype_new, by.x="SNP_ID", by.y="SNP_ID", all = TRUE)


    ggplot(data= m.joined_geno) +
      geom_bar(mapping = aes(x = Chromosome, fill=Homozygous),position="dodge")+
      scale_x_discrete(limits=c(1:10, "unknown", "multiple")) +
      ggtitle(label = "Heterozygosity per Chromosome") +
      xlab(label = "Chromosome #") +
      ylab(label = "Amount of Heterozygosity")

    ```
