---
title: "Sushma_review"
author: "Sushma Ambekar"
date: "3/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Review of Dua Vang's R Assignment

#### Data inspection

Everything ran smooth here and you used appropriate commands to look at the data.


#### Data processing

Everything is easy to understand and runs well.

The transposed files look good, but after merging the two data sets, I noticed that the sample ID row from the transposed maize and teosinte files were retained in the joined_maize and joined_teosinte files, and I'm not sure why that happened, but you can take a look at it.

Just a small thing- I noticed that the code chunks in a few parts of the data processing were not recognized becuase of additional spaces in the document, might be helpful to check for these before submitting to the instructors. 
    

#### Data visualization

All parts of data visualization are very well explained and easy to follow. Plots look great!

For plotting heterozygosity by group, for aesthetics, you could make the group names more visible by using this code instead. It rotates the x axis text by 90 degrees

```{r}
ggplot(data = m.genotype_new) + geom_bar(mapping = aes(x=Group, fill=Homozygous), position="dodge") + theme(axis.text.x = element_text(angle = 90))
```